---
title: "Tonal Languages"
author: "Ali Taqi"
date: "2023-04-28"
output: pdf_document
---

```{r setup, include=FALSE}
# Chunk setup
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
# Load libraries
library(tidyverse)
library(gglm)
```

```{r data, include = FALSE}
# Source scripts
source("data.R")
# Obtain raw data
data_raw <- read_tsv("tone.tsv", show_col_types = F)
# Get wrangled data
tones_complex <- wrangle_data(data = data_raw, scheme = "complex")
```

```{r}
glm_P_sq <- glm(data = tones_complex, 
                   formula = complex_tonal ~ I(dist_equator^2), 
                   family = binomial(link = probit))
summary(glm_P_sq)
```

# Manual Error Analysis

*Idea:* We try to manually find out why equator-close languages are atonal from other variables, and 
add potential variables (for example, colonization history, case of extreme language erasure, etc.)
In Thailand, it was under French Indochina, but it survived? So, need to learn more about each case and find potential sociopolitical factors to explain more variation. This would come from manually studying each case and history.

In data visualization, we found that no tonal language existed farther than 30 latitude. So, we use that as our reference point in order to try to obtain perfect prediction. However, in practice, we might want to use alternative thresholds to really emphasize the outlier status. 

A no-brainer: incorporate climatically outlier zones, where equator doesn't imply humidity, if possible.

```{r}
# Make a copy of the data
outliers <- tones_complex
```

```{r}
table(tones_complex$complex_tonal)
```

```{r, echo = F}
# Threshold Outlier Status
cases_by_threshold <- function(max_latitude){
  total <- nrow(outliers %>% filter(dist_equator < max_latitude) %>% filter(tonality_type == 1))
  total/nrow(tones_complex)
}
if(T){
  data.frame(L = 1:30, Cases = map_dbl(1:30, cases_by_threshold)) %>%
    ggplot() +
    geom_point(aes(x = L, y = Cases))
}
```


```{r}
study <- outliers %>% 
  filter(dist_equator < 30) %>% 
  filter(dist_equator > 20) %>% 
  filter(tonality_type == 1)
```

```{r, fig.height = 14, fig.width = 9}
study %>% 
  ggplot() + 
  geom_histogram(aes(x = family, fill = family), stat = "count") + 
  coord_flip() +
  theme(legend.position = "bottom")
```

```{r}
# Get ggplot coords
world_coordinates <- map_data("world")
# Plot
study %>% 
  ggplot() +
  geom_map(data = world_coordinates, map = world_coordinates, 
           aes(long, lat, map_id = region)) +
  geom_point(aes(x = longitude, y = latitude),  alpha = 0.3, color = "red")
```

